---
title: "Spatial Correlation II"
author: "Adam Peterson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tidyverse)
library(rstap)
```

To induce correlation between business locations, we generate the x and y coordinates from a bivariate normal with correlation $\rho$

```{r}
set.seed(1214)
num_subj <- 5.5E2
sub_df <- data_frame(x = runif(n = num_subj, min = 1, max = 2.0),
                     y = runif(n = num_subj, min = 1, max = 2.0),
                     class = "Subject")
num_bef <- 1.5E2
corr <- .8
v <- rnorm(n = num_bef,mean = 0, sd = 1)
u <- rnorm(n = num_bef,mean = 0, sd = 1)
bef_df <- data_frame(x = pnorm(v)*3,
                     y = pnorm(corr*v + sqrt(1-corr^2)*u)*3,
                     class = 'BEF')
rbind(sub_df,bef_df) %>% ggplot(aes(x=x,y=y,color=class)) + 
    geom_point()
```



Then suppose the form of some measurement on the subjects is:

$$
E[Y_i] = \alpha + Z_i \delta + X_i(\theta)\beta_2\\
X_i(\theta) = \sum_{d \in \mathcal{D}_i} \text{erfc}(\frac{d}{\theta})
$$
Here we use the following 

```{r}
alpha <- 22.5
Z <- rbernoulli(n = num_subj, p = .45)
delta <- -.8
theta <- .5
beta <- 1.2
sigma <- 2.3
```

supposing we use all the data available (no exclusion distance), then 
this results in the following dataset

```{r}
dists <- fields::rdist(as.matrix(sub_df[,1:2]),
                       as.matrix(bef_df[,1:2]))
X <- apply(dists,1,function(x) sum(pracma::erfc(x/theta)))
X_tilde <- (X-mean(X))/sd(X) ## center and scale the exposure effect
y <- alpha + Z*delta + X_tilde*beta + rnorm(n = num_subj, mean = 0, sd = sigma)
hist(y)
```

```{r}
d <- seq(from = 0, to = max(dists), by = 0.01)
X_theta_one <- pracma::erfc(d/theta)
par(mfrow = c(1,2))
plot(d,X_theta_one,type='l',main = "Exposure Decay Function", xlab = "Distance",
     ylab = "Exposure")
hist(X_tilde)
```


Then we model this with rstap in the following manner

```{r}
distance_data <- dists %>% as_data_frame() %>%
    mutate(subj_id = 1:num_subj) %>% 
    gather(contains("V"), key = 'BEF',value = 'Distance') %>% 
    mutate(BEF = 'Fast_Food')

subject_data <- data_frame(subj_id = 1:num_subj,
                           y = y,
                           sex = factor(Z,labels=c("M","F")))

fit <- stap_glm(formula = y ~ sex + stap(Fast_Food),
                subject_data = subject_data,
                distance_data = distance_data,
                family = gaussian(link = 'identity'),
                id_key = 'subj_id',
                prior = normal(location = 0, scale = 5,autoscale = F),
                prior_intercept = normal(location = 25, scale = 5, autoscale = F),
                prior_stap = normal(location = 0, scale = 3, autoscale = F),
                prior_theta = log_normal(location = 0, scale = 1), 
                prior_aux = cauchy(location = 0,scale = 5),
                max_distance = max(dists), chains = 1) ## include all data
```


```{r}
results <- rbind(c(alpha,delta,beta,theta),coef(fit))
rownames(results) <- c("True","Estimated")
results
```

```{r}
fit$ses
```



